---
title: "NYC Jobs"
author: "Roland Gjopalaj"
date: "2023-01-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE)
```

## Overview 

[This dataset contains current job postings available on the City of New York’s official jobs site](http://www.nyc.gov/html/careers/html/search/search.shtml)

Internal postings available to city employees and external postings available to the general public are included.


## Libraries to import 

```{r}
library(tidyr)
library(dplyr)
library(readr)
library(lubridate)
library(ggplot2)
```


## Dataset to import 
The dataset we are going to use to do our studyng.

```{r}
## Import the dataset

jobs = read_csv("JOBS_nyc.csv")
jobs
```

## Cleanin the dataset

If you take a look at the data you will notice that some of the rows are duplicated and that in not efficient so we are going to remove all of theme.

```{r}
## remove the duplicated lines
jobs = jobs[!duplicated(jobs), ]
jobs
```


After that we need to remove the unnecessary data and to keep only the data we want to use.

To facilitate a little bit our work we can change the names of the columns so we don't need to use characters as "`".



```{r}
## Select the necessary data and rename the columns (practical names)

jobs = jobs %>% 
  select(`Job ID`, `Agency`, `Posting Type`, `# Of Positions`, `Title Classification`, `Job Category`, `Full-Time/Part-Time indicator`, `Career Level`, `Salary`, `Work Location`, `Division/Work Unit`, `Posting Date`, `Posting Updated`) %>% 
  rename(Id=`Job ID`, Posting_Type=`Posting Type`, Nr_of_Positions=`# Of Positions`, Classification=`Title Classification`, Job_Category=`Job Category`, FullPart_Time=`Full-Time/Part-Time indicator`, Career_Level=`Career Level`, Work_Location=`Work Location`, Work_Unit=`Division/Work Unit`, Posting_Date=`Posting Date`, Posting_Updated=`Posting Updated`)

 jobs
```

In the dataset we have a column that doesn't contain atomic data so the best choice is to separate it into atomic data so we can use the data properly.


```{r}
## Fix the salary problem (not atomic data)
## separating salary interval in  different columns 

jobs = jobs %>% 
  separate(Salary, into = c("Salary_From", "Salary_To", "Salary_Frequency"), 
         sep = "-", convert = TRUE)

jobs

```
The salary data in this dataset is a little bit a mes as it is in different frequencies.
As we want to do a lot of studies base on the salary we need to create a column (Salary index) that will be in "hours".  (After some researches in America the average of a part-time job is 17 h/week and full-time is 37 h/week and a year has 260 working days)


```{r}
## create a salary index (per hour)

jobs = jobs %>% 
  mutate(Salary_Index = 
           ifelse(Salary_Frequency=="Hourly",
                 (Salary_From+Salary_To)/2,
                 ifelse(FullPart_Time=="P",
                       ## if the work is  part-time
                       ifelse(Salary_Frequency=="Annual",
                             (Salary_From+Salary_To)/(17*260),
                             (Salary_From+Salary_To)/(17)
                       ),
                       ## if the work is  full-time
                       ifelse(Salary_Frequency=="Annual",
                             (Salary_From+Salary_To)/(37*260),
                             (Salary_From+Salary_To)/(37)
                       )
                 )
          )
      )

jobs
```

After that we need to add a new column that will be called ACR that stands for acronym and will be a abbreviation of the agency name (because during the presentation and the visualization the abbreviation will look batter and it wont overlap).

```{r}
##I used this command to ad the ACR for every agency to be more visible on the chart
##jobs = mutate(jobs, ACR = ifelse(Agency=="AGENCY NAME", "ACR", ACR)) 

##Just execute this command to go on 
jobs = read.csv("JOBS_ncyV2.csv")##The dataset after adding the acronym column(ACR)

##select the agencies that have the most number of postings
agencyMore=count(jobs, ACR) %>% 
  arrange(-n) %>% 
  top_n(10) %>% 
  pull(ACR) %>% 
  unlist()

agencyMore = jobs %>% filter(ACR %in% agencyMore)

##visualize it
ggplot(data = agencyMore) +
  geom_bar(mapping = aes(x = ACR))

#####################################################
##select the agencies that have the most number of postings
agencyLess=count(jobs, ACR) %>% 
  arrange(n)  %>% 
  pull(ACR) %>% 
  unlist()

##another way to select the first 10
agencyLess=agencyLess[0:10]

agencyLess = jobs %>% filter(ACR %in% agencyLess)

ggplot(data = agencyLess) +
  geom_bar(mapping = aes(x = ACR))

```

```{r}
jobs = read.csv("JOBS_ncyV2.csv")

##fix the date(convert it from string to date)
jobs = jobs %>% 
  separate(Posting_Date, c("month", "day", "year"), sep = "/") %>% 
  mutate(Posting_Date = make_date(year, month, day)) %>% 
  select(Id:Work_Unit, Posting_Date, Posting_Updated:ACR)
jobs = jobs %>% 
  separate(Posting_Updated, c("month", "day", "year"), sep = "/") %>% 
  mutate(Posting_Updated = make_date(year, month, day)) %>% 
  select(Id:Posting_Date, Posting_Updated, Salary_Index, ACR)


```


```{r}

##change some values for better visibility 
jobs["Career_Level"][jobs["Career_Level"] == "Experienced (non-manager)"] = "EXP-NO-Manager"


## show the salaries in base of the career level
## and with the help of the colors we can see the level of the difficulty 
## we can also filter the time (from a given year to the day of 17/01/2013)
show = function(posting_type, myDateFrom) {
  jobs[!is.na(jobs$Career_Level), ] %>% ##remove the NA elements 
    filter(Salary_From>0, Posting_Type==posting_type, Posting_Date > myDateFrom) %>% 
    ggplot(aes(Career_Level, Salary_Index, colour = Classification, size=Nr_of_Positions)) +
    geom_point(alpha = 0.4, na.rm = TRUE) +   
    theme_minimal() 
}



show("External", as.Date(paste(2015,"-", 1, "-",1, sep = "")))
show("External", as.Date(paste(2022,"-", 1, "-",1, sep = "")))
show("Internal", as.Date(paste(2023,"-", 1, "-",1, sep = "")))


```

```{r}
## Top 10 most frequent Job_category
topCategory = jobs %>% 
  count(Job_Category) %>% 
  pull(Job_Category) %>% 
  unlist()
topCategory = topCategory[0:5]

ggplot(data = filter(jobs, Job_Category %in% topCategory)) +
  geom_bar(mapping = aes(x = Job_Category))+
  scale_x_discrete(guide = guide_axis(n.dodge=5))


```





```{r}
## posts durind the years based on the number of opened positions, divided by the classification group, and with some salary details.
jobs %>% 
  ggplot(aes(Posting_Date, Nr_of_Positions, colour = Classification, size= Salary_Index))+
  geom_point(alpha = 0.5, na.rm = TRUE)+
   scale_x_date(date_labels = "%Y")

```



```{r}

##create  a column difficulty(numeric) that will be the numeric interpretation of classification

## First we remove the "-" symbol to avoid problems with separating
jobs["Classification"][jobs["Classification"] == "Non-Competitive-5"] = "NonCompetitive-5"

##Here we want to create a model of how the difficulty of a job goes based on the salary


jobsDiff=jobs %>% 
  separate(Classification, c("Classification","Difficulty"), sep = "-")

mod = lm( Difficulty ~ Posting_Date, data = jobsDiff)
##summary(mod)


jobsDiff %>% 
  ggplot(aes(Posting_Date,Difficulty))+
  geom_point(na.rm = TRUE)+
  geom_abline(intercept = mod$coefficients[1], 
              slope = mod$coefficients[2], 
              color = "red")

```














